---
title: "Initial Analysis"
author: "Dwipam, Neelam"
output: html_document
---

##Variable description:
FID:ROW_ID  
ST_CASE:CASE_ID  
LATITUDE:LATITUDE (LOCATION)  
LONGITUDE:LONGITUDE (LOCATION)  
PERMVIT: Number of Persons in Motor Vehicles in Transport  
PERSONS: Total number of fatalities  
MONTH: Month of crash  
DAY_WEEK: Weekday of crash  
HOUR: Hour of crash  
ROUTE: Route Signing  
1 Interstate  
2 U.S. Highway  
3 State Highway  
4 County Road  
5 Local Street – Township  
6 Local Street – Municipality  
7 Local Street – Frontage Road (Since 1994)  
8 Other  
9 Unknown  
SP_JUR:  
HARM_EV:  
RELICT1:  
RELICT2:  
TYP_INT:  
WRK_ZONE:  
REL_ROAD:  
LGT_COND:  
WEATHER:  
FATALS:  
STATE:  
A_RU:  
A_INTER:  
A_TOD:  
A_DOW:  
A_CT:
A_MC:
A_SPCRA:
A_PRED:
A_ROLL:
A_POLPUR:
A_POSBAC:
A_D15_20:
A_D65PLS:
A_D21_24:
A_HR:
A_DIST:
A_DROWSY:

```{r,message = FALSE}
library(dplyr)
county = read.csv('us_county.csv') #Received from prof. D Ficklin from geology department
fatalities = read.csv('final_data.csv')
county = county %>% select(STATE_NAME,STATE_FIPS)
county = county[!duplicated(county$STATE_NAME),]
fatalities = merge(fatalities,county,by.x = "OBJECT_ID",by.y = "STATE_FIPS",all.x =FALSE)
```
##Exploratory Data Analysis:  
Now we have the data in the desired form with the State_name and their associated values. Lets do some exploratory data analysis:
```{r}
library(ggplot2)
ggplot(fatalities) + aes(x = STATE_NAME) + geom_bar(stat = 'count') + coord_flip() + ylab("Total Number of fatalities accident") + xlab("State Name") + ggtitle("Fatalities accident by State")
```
It seems that Texas, has the highest number of fatalities accident in U.S. Also these states has higher number of population in U.S. Let's normalize by the population.  
#####Population data is collected from 
```{r}

```
